Stateless Model Checking with Data-Race Preemption Points

### 2016-07-15 oopsla-16/implementation.tex
Improve the prose.
mando revisions regarding chess, fasttrack, djit, and seqcst

We chose \landslide~\cite{landslide} as our MC tool due to its ability to trace execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
\landslide~implements DPOR \cite{dpor},
state space estimation \cite{estimation}, and a hybrid
lockset/Limited HB
data-race analysis \cite{hybriddatarace}.
\revision{It can optionally replace DPOR with ICB \cite{chess-icb},
for which it uses Bounded Partial Order Reduction \cite{bpor} for a similar reduction,
%although we leave incorporating ICB with Iterative Deepening to future work (\sect{\ref{sec:future}})}.
though \quicksand~does not employ this feature (\sect{\ref{sec:future}})}.
It avoids state space cycles
%(e.g. ad-hoc synchronization with {\tt yield} or {\tt xchg} loops)
\revision{(e.g. spin
%or {\tt yield}
loops)}~with
a heuristic similar to Fair-Bounded Search \cite{bpor}.
% joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
Its bug-detection metrics are assertion failure, deadlock, segfault, use-after-free \cite{valgrind}, and (heuristically) infinite loops.
%a heuristic infinite loop check.

@@

\quicksand~is an independent program that wraps the execution of several \landslide~MC instances.
The implementation is roughly 3000 lines of C.
%It uses a thread pool to schedule the available state spaces,
%sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
%Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
The interface with the MC has two parts. %, which any similar MC could implement, has two parts.
First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
% can lose this line due to space
plus other MC-specific options such as our modifications to \landslide~for testing mutexes. %mutex-testing mode,
%passed as an argument to \landslide.
Then, a dedicated \quicksand~thread communicates with the MC process via message-passing. %on a FIFO pipe.
%\landslide~messages \quicksand~
The MC messages after testing each interleaving to report updated progress and ETA
and whenever a new data-race candidate or bug is found.
\quicksand~in turn replies whether to resume/suspend (due to too high ETA), or quit (due to timeout).
We suspend jobs simply by making the MC wait on a message-passing reply.
Should \quicksand~later re-schedule a suspended job, we send a message to continue,
resuming it where it left off.
%otherwise, we resume it only after time runs out, causing it to exit.

>>>>

We chose \landslide~\cite{landslide} as our MC tool due to its ability to trace execution at the granularity of individual instructions and memory accesses, which dynamic data-race detection requires.
\landslide~implements DPOR \cite{dpor},
state space estimation \cite{estimation},
\revisionx{the DJIT+ HB data-race analysis \cite{djit,fasttrack},}
and a hybrid
lockset/Limited HB
data-race analysis \cite{hybriddatarace}.
\revision{It can optionally replace DPOR with ICB \cite{chess-icb},
for which it uses Bounded Partial Order Reduction \cite{bpor} for a similar reduction,
%although we leave incorporating ICB with Iterative Deepening to future work (\sect{\ref{sec:future}})}.
though \quicksand~does not employ this feature (\sect{\ref{sec:future}})}.
It avoids state space cycles
%(e.g. ad-hoc synchronization with {\tt yield} or {\tt xchg} loops)
\revision{(e.g. spin
%or {\tt yield}
loops)}~with
a heuristic similar to Fair-Bounded Search \cite{bpor}.
% joshua wants "segfault" to be "memory access error (i.e., segmentation fault, or bus error)"
Its bug-detection metrics are assertion failure, deadlock, segfault, use-after-free \cite{valgrind}, and (heuristically) infinite loops.
%a heuristic infinite loop check.

@@

\quicksand~is an independent program that wraps the execution of several \landslide~MC instances.
The implementation is roughly 3000 lines of C.
%It uses a thread pool to schedule the available state spaces,
%sorting such jobs according to their status among a running queue, pending queue, and suspended queue.
%Jobs are further prioritized by number of PPs, ETA, and whether they include data-race PPs.
The interface with the MC has two parts. %, which any similar MC could implement, has two parts.
First, when starting each job, \quicksand~creates a configuration file declaring which PPs to use,
% can lose this line due to space
plus other MC-specific options such as our modifications to \landslide~for testing mutexes. %mutex-testing mode,
%passed as an argument to \landslide.
Then, a dedicated \quicksand~thread communicates with the MC process via message-passing. %on a FIFO pipe.
%\landslide~messages \quicksand~
The MC messages after testing each interleaving to report updated progress and ETA
and whenever a new data-race candidate or bug is found.
\quicksand~in turn replies whether to resume/suspend (due to too high ETA), or quit (due to timeout).
We suspend jobs simply by making the MC wait on a message-passing reply.
Should \quicksand~later re-schedule a suspended job, it sends a message to continue,
resuming the job where it left off.
%otherwise, we resume it only after time runs out, causing it to exit.

### 2016-07-15 oopsla-16/intro.tex
mando revisions regarding chess, fasttrack, djit, and seqcst

However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
For example, \textsc{CHESS} \cite{chess} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
%
On the other hand, SPIN \cite{spin} and Inspect \cite{inspect}
are able to preempt threads before any shared memory access.
Such fine granularity would automatically check each data race for the possibility of failure, but risks timing out before the state space can be completed. %makes full state space completion intractable for even modestly-sized tests.
Choosing preemption points is a tradeoff between schedule coverage and feasibility of completion,
but even with state-of-the-art reduction techniques,
\revision{fixing the degree of coverage}~in advance necessarily leaves some tests unaffordably large \cite{parrot,mcr}.
%This work shows how to avoid making that tradeoff decision in advance.

@@

Our contributions are as follows:
\begin{enumerate}
	\item {\em Iterative Deepening}, a new \revision{algorithm}~for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation;
	\item A proof of convergence, showing that should it be possible in the given CPU budget,
		fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules;
	\item A new tactic for eliminating one class of false-positive data race candidates,
		which cannot soundly be used in a single-pass analysis,
		but which we prove correct when used with Iterative Deepening;
	\item A large evaluation in which \quicksand~compares favourably to stand-alone data-race detection and stateless model checking approaches, finding new bugs that would be missed by either alone.
\end{enumerate}

>>>>

However, all stateless model checkers to date are bound by a fixed set of {\em preemption points}: code locations that define the granularity at which threads interleave.
For example, \textsc{CHESS} \cite{chess} \revisionx{by default} preempts only on synchronization operations and library calls, which can miss lock-free shared memory races.
%
On the other hand, SPIN \cite{spin} %and Inspect \cite{inspect}
%are able to preempt
preempts threads before any shared memory access.
Such fine granularity would automatically check each data race for the possibility of failure, but risks timing out before the state space can be completed. %makes full state space completion intractable for even modestly-sized tests.
\revisionx{Some tools, such as CHESS and Inspect \cite{inspect} can strike a middle ground using compiler instrumentation to statically add preemption points on memory accesses.
Nevertheless,}
choosing preemption points is a tradeoff between schedule coverage and feasibility of completion:
%but
even with state-of-the-art reduction techniques,
\revision{fixing the degree of coverage}~in advance necessarily leaves some tests unaffordably large \cite{parrot,mcr}.
%This work shows how to avoid making that tradeoff decision in advance.

@@

Our contributions are as follows:
\begin{enumerate}
	\item {\em Iterative Deepening}, a new \revision{algorithm}~for combining data-race analysis with stateless model checking, and \quicksand, an open-source implementation;
	\item A proof of convergence \revisionx{for sequentially-consistent memory models}, showing that should it be possible in the given CPU budget,
		fully testing every discovered data-race preemption point is equivalent to testing all possible thread schedules;
	\item A new tactic for eliminating one class of false-positive data race candidates,
		which cannot soundly be used in a single-pass analysis,
		but which we prove correct when used with Iterative Deepening;
	\item A large evaluation in which \quicksand~compares favourably to stand-alone data-race detection and stateless model checking approaches, finding new bugs that would be missed by either alone.
\end{enumerate}

### 2016-07-15 oopsla-16/overview.tex
Last of garth's commence
mando revisions regarding chess, fasttrack, djit, and seqcst

\subsection{Stateless Model Checking}

@@

\revision{
	{\bf Variants of Happens-Before.}
	Most prior work focuses on {\em Happens-Before} (HB) \cite{lamport-clocks} as the order relation between accesses.
\cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:
it cannot identify access pairs separated by an unrelated lock operation which could race in an alternate interleaving,
as shown in the example program in Figure~\ref{fig:hb-example}(a).
We call such unreported access pairs {\em false negatives}.
\cite{hybriddatarace} introduces the {\em Limited HB} relation,
which will report such potential races
by considering only blocking operations like {\tt cond\_wait} to enforce the order.
However, consider the similar program in Figure~\ref{fig:hb-example}(b),
in which the access pair ceases to exist in the alternate interleaving.
Limited HB will report all potential races, avoiding false negatives \cite{tsan},
but at the cost of necessarily reporting some such {\em false positives}.

@@

Stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx}.
%However,
This work incorporates data-race analysis in an internal feedback loop,
using model checking to automatically test each potential race
and report only directly observed failures to the user.
%Our convergence theorem requires identifying all potential data-races to provide total verification (\sect{\ref{sec:totalverif}})
%% TODO CAMREADY: cite our soundness proofs tech report in this paragraph as well
Hence, we employ Limited HB in this paper},
accepting some overhead from false positives for the sake of total verification (\sect{\ref{sec:totalverif}}).

>>>>

\subsection{Stateless Model Checking}
\label{sec:overview-sssmc}

@@

	{\bf Variants of Happens-Before.}
	Most prior work focuses on {\em Happens-Before} (HB) \cite{lamport-clocks,djit,fasttrack} as the order relation between accesses.
\cite{predictive-dr} and \cite{hybriddatarace} identify a problem with this approach:
it cannot identify access pairs separated by an unrelated lock operation which could race in an alternate interleaving,
as shown in the example program in Figure~\ref{fig:hb-example}(a).
We call such unreported access pairs {\em false negatives}.
\cite{hybriddatarace} introduces the {\em Limited HB} relation,
which will report such potential races
by considering only blocking operations like {\tt cond\_wait} to enforce the order.
However, consider the similar program in Figure~\ref{fig:hb-example}(b),
in which the access pair ceases to exist in the alternate interleaving.
Limited HB will report all potential races, avoiding false negatives \cite{tsan},
but at the cost of necessarily reporting some such {\em false positives}.

@@

Stand-alone data-race analyses must avoid inundating the user with false alarms \cite{racerx}.
%However,
This work incorporates data-race analysis in an internal feedback loop,
using model checking to automatically test each potential race
and report only directly observed failures to the user.
%Our convergence theorem requires identifying all potential data-races to provide total verification (\sect{\ref{sec:totalverif}})
%% TODO CAMREADY: cite our soundness proofs tech report in this paragraph as well
%Hence, we employ Limited HB in this paper},
%accepting some overhead from false positives for the sake of total verification (\sect{\ref{sec:totalverif}}).
\revisionx{Hence, we can accept some overhead from Limited HB's false positives for the sake of finding data-race candidates more quickly.
In \sect{\ref{sec:eval}} we will evaluate how HB and Limited HB each influence \quicksand's bug-finding and verification speed.
}

### 2016-07-15 oopsla-16/related.tex
mando revisions regarding chess, fasttrack, djit, and seqcst

{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, MaceMC \cite{macemc}, MoDist \cite{modist}, ETA \cite{dbug-retreat}, and Concuerror \cite{optimal-dpor},
limit thread communication to a message-passing API to more effectively test distributed systems.
This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
Nevertheless, Iterative Deepening is applicable to these tools.

@@

%Too many related projects to list have made contributions to the
Many advances have been made on the false-positive potential data race problem since it was first introduced in \cite{eraser}.
\cite{hybriddatarace} and \cite{tsan}
%combine the lockset and happens-before analyses into a hybrid technique,
\revision{introduce Limited Happens-Before}, %(\sect{\ref{sec:overview-dr}})},
which we employ,
\revision{while \cite{predictive-dr} introduces the Causally-Precedes relation for a sound yet incomplete analysis (\sect{\ref{sec:overview-dr}})}.

\revision{Several tools \cite{droidracer,cafa,unnamed-android} have recently emerged to target Android applications},
%DroidRacer \cite{droidracer} and CAFA \cite{cafa} target
%%event-driven
%Android applications,
using domain-specific heuristics (orthogonal to our method) to reduce false positives. % cut for space?
Like \landslide, DataCollider \cite{datacollider} \revision{finds data races in}~kernel code.
IFRit \cite{ifrit}
improves the performance of HB using an interference analysis,
which could allow future work to avoid tracing every memory access.
%although in our context, we cannot admit false negatives (\sect{\ref{sec:totalverif}}).
% No, IDGAF about pure happens before.
%FastTrack \cite{fasttrack} optimizes the performance of pure happens-

Closer to our work, replay analysis \cite{recordreplaydrs} also suppresses false positives by testing multiple thread interleavings.
%after finding data race candidates.
This work compares the program states immediately after the access pair for differences,
preferring to err on the side of false positives.
RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as we do,
although it uses random schedule fuzzing rather than stateless MC.
While this technique can classify malloc-recycle candidates as false positives (\sect{\ref{sec:recycle}}),
they require replaying the threads in a new interleaving.
Moreover, \cite{portend} argues that accurate classification may require many re-executions,
%according to many pre- and post-race sequences,
which is tantamount to adding a new state space in \quicksand.
Our proof in \sect{\ref{sec:recycle}} allows us to eliminate this special case with no additional replay beyond what DPOR already requires.

>>>>

{\bf Message-passing.} Other stateless model checkers, such as SAMC \cite{samc}, MaceMC \cite{macemc}, MoDist \cite{modist}, ETA \cite{dbug-retreat}, and Concuerror \cite{optimal-dpor},
limit thread communication to a message-passing API to more effectively test distributed systems.
This eliminates the need for data-race analysis, but restricts the class of programs that can be tested.
%Nevertheless, Iterative Deepening is applicable to these tools.

@@

\revisionx{{\bf Happens-Before.}}
%Too many related projects to list have made contributions to the
Many advances have been made on the false-positive potential data race problem since it was first introduced in \cite{eraser}.
\cite{hybriddatarace} and \cite{tsan}
%combine the lockset and happens-before analyses into a hybrid technique,
\revision{introduce Limited Happens-Before}, %(\sect{\ref{sec:overview-dr}})},
which
\revisionx{\quicksand~uses to achieve its best bug-finding result.
Other tools such as DJIT+ \cite{djit} and FastTrack \cite{fasttrack} opt for the precise Happens-Before analysis,
which produces no false positives but is prone to more false negatives as shown in \sect{\ref{sec:overview-dr}}.
FastTrack, the most recent development in this approach,
optimizes the representation of variable write vector clocks for performance,
which guarantees to detect at least the first race on each variable.
However, as we are interested in classifying data race candidates as {\em benign} or {\em buggy},
this optimization would be unsound for total verification if the first race on a variable were benign.
Hence, we implement precise HB as defined by DJIT+.
Finally, \cite{predictive-dr} recently introduced the Causally-Precedes relation for a precise analysis which covers some common false negative cases
such as our example in Figure~\ref{fig:hb-example}(a).
}
%\revision{while \cite{predictive-dr} introduces the Causally-Precedes relation for a sound yet incomplete analysis (\sect{\ref{sec:overview-dr}})}.

\revisionx{{\bf Other domains.}}
\revision{Several tools \cite{droidracer,cafa,unnamed-android} have recently emerged to target Android applications},
%DroidRacer \cite{droidracer} and CAFA \cite{cafa} target
%%event-driven
%Android applications,
using domain-specific heuristics (orthogonal to our method) to reduce false positives. % cut for space?
Like \landslide, DataCollider \cite{datacollider} \revision{finds data races in}~kernel code.
IFRit \cite{ifrit}
improves the performance of HB using an interference analysis,
which could allow future work to avoid tracing every memory access.
%although in our context, we cannot admit false negatives (\sect{\ref{sec:totalverif}}).
%FastTrack \cite{fasttrack} optimizes the performance of pure happens-

\revisionx{{\bf Replay analysis.}}
Closer to our work, replay analysis \cite{recordreplaydrs} also suppresses false positives by testing multiple thread interleavings.
%after finding data race candidates.
This work compares the program states immediately after the access pair for differences,
preferring to err on the side of false positives.
RaceFuzzer \cite{racefuzzer} avoids false positives by requiring an actual failure be exhibited, as we do,
although it uses random schedule fuzzing rather than stateless MC.
While this technique can classify malloc-recycle candidates as false positives (\sect{\ref{sec:recycle}}),
they require replaying the threads in a new interleaving.
Moreover, \cite{portend} argues that accurate classification may require many re-executions,
%according to many pre- and post-race sequences,
which is tantamount to adding a new state space in \quicksand.
Our proof in \sect{\ref{sec:recycle}} allows us to eliminate this special case with no additional replay beyond what DPOR already requires.

### 2016-07-15 oopsla-16/soundness.tex
Improve the prose.
mando revisions regarding chess, fasttrack, djit, and seqcst

@@

\revision{This class of false positive is unique to heap-allocated memory among all ways threads could communicate.
By contrast, global memory has unlimited lifetime,
and message-passing primitives enforce an ordering which precludes the race. %(even under Limited HB).
Finally, note that as long as concurrent {\tt malloc} is implemented with an internal lock,
these false positives are of concern only under a Limited HB analysis (\sect{\ref{sec:overview-dr}}).
Nevertheless,
%our previous proof's
Theorem~\ref{thm:convergence}'s need for the Limited HB definition justifies our choice of it over Pure HB.
% TODO CAMREADY: Cite the tech report proofs, which I think call out the use of limited HB in the proof more explicitly.
}

>>>>

\revisionx{These proofs are built on a DPOR algorithm definition which assumes sequentially-consistent memory hardware.
%
% TODO: Aaargh, can't think of how to actually justify this.
%We believe this assumption is reasonable for the scope of this paper because the main property we require of DPOR is that 
A recent paper extended DPOR to support weaker memory models \cite{tsopso}, so our proofs could likewise be extended with that version of DPOR in future work.
We also assume the Limited HB definition for the data-race analysis, leaving the case for precise HB to future work.
}

@@

\revision{This class of false positive is unique to heap-allocated memory among all ways threads could communicate.
By contrast, global memory has unlimited lifetime,
and message-passing primitives enforce an ordering which precludes the race. %(even under Limited HB).
Finally, note that as long as concurrent {\tt malloc} is implemented with an internal lock,
these false positives are of concern only under a Limited HB analysis (\sect{\ref{sec:overview-dr}}).
Nevertheless,
%our previous proof's
Theorem~\ref{thm:convergence}'s need for the Limited HB definition justifies our choice of it over full HB.
% TODO CAMREADY: Cite the tech report proofs, which I think call out the use of limited HB in the proof more explicitly.
}
